---
title: "Working with rasters in the terra package"
author: "Anna Stroh"
output:
  html_document: default
  pdf_document: default
date: "2024-09-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# keep code just for knitting - to be copied into console
#pandoc::with_pandoc_version(
 # version = '2.9.1',
  #rmarkdown::render('input.Rmd')
#)

```

This vignette focuses primarily on rasters, and aims to demonstrating 
fundamental raster operations needed to pre-process spatiotemporal rasters for 
further use, i.e. modelling. This vignette includes code to:

* load, 
* visualise,
* manipulate, and 
* extract data into csv formats 

from spatiotemporal raster files using the R package `terra`.  

## _Relevant packages_
```{r}

library(terra) # raster work
library(sf) # vector work

library(ggplot2)
library(tidyterra) # for SpatRaster/SpatVector plotting 
library(viridis) # viridis colour-scale

library(dplyr) # data wrangling
library(tidyr)

```

## _Load raster data from ERDDAP servers_

A spatio-temporal raster is inherently multi-layered, as it stores data in each 
layer corresponding to a time stamp. This is often referred to a raster brick
or data cube in other raster-specialised packages (`stars`). `terra` treats both 
spatial and spatiotemporal rasters as `SpatRaster`. 

Spatiotemporal grid data can be accessed within R using the `rerddap` package (as
demonstrated in this vignette). ERDDAP servers archive environmental, 
meteorological, oceanographic and biological data. 

For this vignette, we use the Marine Institute Ireland 
[ERDDAP server](https://erddap.marine.ie/erddap/) to load sea surface 
temporature (SST). 
This dataset provides monthly mean sea surface temperature from 2016-current in 
Irish Western waters. Suppose this is the available raster file given for the
research project.

The loaded data often comes in form of a NetCDF file (filename.nc), which is 
comparable to an array in R. NetCDF files can be conveniently read in
using `rast()`. 

This code chunk is based Dr Cóilín Minto's code.

```{r rerddap data}
#install.packages("rerddap")
library(rerddap)

urlBase <- "https://erddap.marine.ie/erddap/" 
parameter <- "sea_surface_temperature"
times <- c("2016-05-15T00:00:00Z", "2024-08-15T00:00:00Z")
lats <- c(51, 54.5) # Western Ireland
lons <- c(-13, -9) 
dataInfo <- rerddap::info("IMI_Model_Stats", url = urlBase)

dat <- griddap(dataInfo,
               longitude = lons,
               latitude = lats,
               time = times,
               fields = parameter,
               url = urlBase,
               read = FALSE,
               store = disk()) # nc file

file.size(dat$summary$filename)/1e6 ## in MB

## Accessing nc file and read in as terra object
rast <- rast(dat$summary$filename) 

## Inspect spatiotemporal raster 
rast

# new: time dimension
head(time(rast)) # get information within time dimension

# more advanced: multilayer raster
nlyr(rast)

## Visualisation 
plot(rast)

```

## _Cropping a raster to a desired spatial extent_ 

Suppose, a raster file needs to be cropped to a custom spatial extent or object,
i.e. an MPA. In `terra`, this is done using `crop()`.

A commonly used application of `crop()`, is cropping a raster to the extent of 
a spatial vector. A vector can be in the shape of points, lines or polygons. 
Each of these data structures consist of pairs of coordinates 
(i.e. latitude, longitude). 

The MPA here is assumed to be polygon (a set of closed lines). 

In `terra`, vectors are created using `vect()`, which creates a 
`SpatVector` object.  

```{r create SpatVector}

## Create MPA SpatVector based on extent
e <- ext(-12.8, -11.8, 52.5, 53.7) # define vector extent
mpa <- vect(e) # create vector (SpatVector)
crs(mpa) <- 'wgs84'

mpa 
plot(mpa)

## Plot MPA with raster
plot(rast[[1]], main = "Original raster with MPA")
plot(mpa, add = TRUE, col = "transparent", border = "white", lwd = 3.5)

## Crop temperature SpatRaster to MPA SpatVector
mpa_rast <- crop(rast, mpa)  
mpa_rast # produces SpatRaster

rm(list = c("rast")) # remove original rast from environment

```

## _Short interlude on plotting `terra` objects in ggplot2_

See here code to map multi-layer `SpatRasters` in `tidyterra`. For 
`SpatVectors` see the end on the vignette.

Note: Personally, I find `terra::plot` advantageous at times, as it automatically
provides single scales for each layer. `tidyterra::geom_spatraster` appears to 
mold all scales in a single joint scale. I think this may result in a less ideal 
mapping of rasters with fine scale differences from one layer to the next.

```{r ggplot terra}

# ggplot: single layer mapping
ggplot() + 
  # if displaying a single layer, use "aes(fill = <layer name>)"
  geom_spatraster(data = mpa_rast, aes(fill = sea_surface_temperature_1)) +
  scale_fill_viridis_c(labels = scales::label_number(suffix = "º"))


# ggplot: multi-layer mapping
 ggplot() +
  geom_spatraster(data = mpa_rast[[1:4]]) # reduce nlyr for plotting + 
  scale_fill_viridis_c(labels = scales::label_number(suffix = "º")) +
  # if displaying multiple layers, use "facet_wrap( ~ lyr)"
  facet_wrap(~ lyr, # no substitute for layer names necessary - use "~lyr"
             ncol = 2, # customise layout
             nrow = 2)
 
 #in comparison
 plot(mpa_rast[[1:4]])

```

## _Subsetting and aggregating rasters_

Subsetting raster layers and time dimension in R is done using double square
brackets `[[ ]]`.

Subsetting layers is usually aided by using numerical indices that match the 
order of layers. The same goes for subsetting the time dimension.  

To subset values (i.e. temperature) within the raster, we need to 'clamp' 
the values. This is not relevant to the current analysis but an example is provided.

```{r Subset terra rasters}

## Subset SpatRaster layers

# to single layer
subset1 <- mpa_rast[[1]] # [[ layer index ]]

# to first three layers
subset2 <- mpa_rast[[1:3]] # [[ layer indexes ]]
nlyr(subset2)

# custom number of layer
subset3 <- mpa_rast[[c(1, 3:6, 9)]] # [[ c(layer indexes) ]]


## Subset raster time dimension
head(time(mpa_rast))
class(time(mpa_rast))

#install.packages("lubridate")
library(lubridate) # to work with dates

time(mpa_rast) <- ymd(time(mpa_rast)) # transform time dimension to year-month-day (Date) format
dates_of_interest <- time(mpa_rast)[1:10] # matching index

time_example <- mpa_rast[[ time(mpa_rast) %in% dates_of_interest ]] # subset time and assign to new raster

## Subset raster values (clamping)

# Clamp values from mean-sd to mean+sd temperatures 

start <- mean(values(mpa_rast)) - sd(values(mpa_rast))
end <- mean(values(mpa_rast)) + sd(values(mpa_rast))

clamp_example <- clamp(mpa_rast, 
                       start, # minimum of desired range
                       end, # maximum of desired range
                       # fill cells w/ values outside of range w/ extreme values
                       values = TRUE)  
#clamp_example

# Compare raster values before and after clamping
par(mfrow = c(1, 2))
plot(mpa_rast, 2, main = "Original raster") 
plot(clamp_example, 2, main = "After clamping")


```

Now, to an applied example.

Suppose that, as part of a long-term monitoring programme, we have sampling 
stations within the MPA for which fish data has been collected between 2019 to 
2024. Our aim is to calculate mean summer SST for these stations, and extract 
them afterwards. 

In this bit, I will also show how `terra` can be used in connection with the 
`sf` (package)[https://r-spatial.github.io/sf/]. In short, the `sf` package 
specialises in the use, manipulation, calculation of vector data. Here, it is
used to create a spatial vector of points within the MPA, representing sampling
stations.

Note: The number of sampling stations has been intentionally kept low to allow 
efficient computation on all devices. 

``` {r make data points}

#install.packages("sf")
library(sf)

## Make data frame with station coordinates
mpa_sf <- st_as_sf(mpa) # convert SpatVector to sf object

# station IDs
IDs <- data.frame(ID = 1:10)

# randomly sample 10 stations 
set.seed(5)
station_coords <- st_sample(mpa_sf, 10, type = "random")
stations_sf <- st_sf(IDs, geometry = station_coords)

st_crs(mpa_sf) == st_crs(stations_sf)

# plot out coords 
ggplot() +
  geom_sf(data = mpa_sf) +
  geom_sf(data = stations_sf)

# Bind sampling dates
# define station sampling dates => using raster time stamps for simplicity
dates <- data.frame(r_time = time(mpa_rast)) |>
  mutate(year = year(r_time),
         month = month(r_time)) |>
  filter(year > 2018,
         month >= 5 & month <= 8) 
sampling_dates <- dates$r_time

stations_container <- list()

for (i in 1:nrow(stations_sf)) {
  
  # extract coordinates
  coords <- st_coordinates(stations_sf[i, ]) # access coordinates
  lon <- coords[1]
  lat <- coords[2]
  
  # create df with coords and sampling dates (raster time stamps for simplicity)
  stations_time <- data.frame(
    StationID = stations_sf$ID[i],
    Long = lon, 
    Lat = lat,
    SamplingDate = sampling_dates
  )
  
  # bind loop df to container
  stations_container[[i]] <- stations_time
}

mpa_stations <- do.call(rbind, stations_container) # make final df
head(mpa_stations)

# make SpatVector from dataframe
mpa_stations_vec <- vect(mpa_stations, geom = c("Long", "Lat"), crs = "wgs84")
dim(mpa_stations_vec)

```

With the sampling stations being set-up, we need to prepare the 
raster data for extraction. Alternatively, we can extract the raw environmental
data here, but when working with big raster files, it can be beneficial to 
calculate the needed values within the raster before extraction.

So, to extract the mean summer SST from the raster, we need to

* subset the cropped raster to the range of sampling dates,
* subset the previous subset to both summer months, and
* aggregate raster subsets to the mean seasonal sea surface 
temperature. 

We define summer months as ranging from May to August. 

Here, we are work with temporal aggregation. `tapp()` accesses raster 
layers indexed by a time index (i.e. years), and aggregates 
these layers over a given function (i.e. mean).

```{r Subset to summer}

## Subset raster time dimension to sampling dates (also summer time)
r_subset <- mpa_rast[[ time(mpa_rast) %in%  sampling_dates ]]

## Aggregate for yearly summer value 
# seasonal mean temperature for each year
summer_agg <- tapp(r_subset, "years", fun = mean)
summer_agg # note change in layer names and time dimension

#terra::units(summer_agg) <- "C"
#names(summer_agg) <- gsub("y_", "summer_", names(summer_agg))
plot(summer_agg)

```
# Part 4: __Raster extraction__

The last part of this vignette demonstrates how to extract raster values for a 
given spatial geometry (point here). This is particularly useful for
linking raster-based environmental data to biological data at sampling locations 
(i.e. stations). 

Here, we aim to link mean summer SST to the MPA sampling stations. 

Raster data for a given object is extracted using the `extract()` function.

To bind the extracted data, the most straightforward way is to include the 
argument `bind=TRUE` which cbinds the extracted raster values to the 
`SpatVector` of the sampling stations. 

We can then transform the `SpatVector` with the extracted value into a 
data frame to further shape it into a suitable format.  

``` {r extract data points}

## Extract temperature data
# reduce to coordinates make SpatVector from dataframe
mpa_stations_vec <- mpa_stations |>
  select(StationID, Long, Lat) |>
  distinct() |>
  vect(geom = c("Long", "Lat"), crs = "wgs84") # make df a SpatVector
dim(mpa_stations_vec)

sst_extr <- terra::extract(summer_agg, # SpatRaster
                           mpa_stations_vec, # SpatVector
                           # returns value from the cell the coords fall into
                           method = "simple",
                           # binds raster values to vector
                           bind = TRUE) # do not return cell ID

head(sst_extr) 

# make dataframe
sst_extr_df <- terra::as.data.frame(sst_extr) # wide format
head(sst_extr_df)

## Reshape and clean-up

#install.packages("reshape2")
# wide to long
sst_extr_melt <- reshape2::melt(sst_extr_df, 
                                id = c("StationID"))
head(sst_extr_melt)

station_sst <- sst_extr_melt |>
  
  # rename layers
  dplyr::rename(Year = variable,
                MSST = value) |>
  
  # clean up year variable
  mutate(Year = gsub("y_", "", Year)) |>
  
  # merge in station sampling date and locations
  merge(mpa_stations) |>
  
  # change order of columns
  relocate(SamplingDate, .before = Year)

head(station_sst) # FINAL DATA FRAME READY FOR EXPORT


## Plot final output
st_vec <- vect(station_sst, geom = c("Long", "Lat"), crs = "wgs84")
head(st_vec)
st_vec

ggplot() +
  geom_spatvector(data = st_vec, aes(colour = MSST)) + # plot SpatVector
  scale_colour_viridis_c(
    labels = scales::label_number(suffix = "º")) +
  geom_spatvector(data = mpa, fill = NA, colour = "black") +
  facet_wrap( ~ Year) +
  labs(colour = "Mean\nSummer\nSST") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 90))



```
# Extra: __Raster animation__

This part is not compulsory for raster analysis, but demonstrates a fun way to
visualise raster data. `terra` has an in-built function to sequentially plot 
raster layers, which can then be exported as a video or gif file: an animation of
the raster. 

Note: I am intentionally using the aggregated raster with regards to rendering 
processes. 

```{r raster animation}

## Make animation
terra::animate(summer_agg, n = 1)

## Save animation as gif
#install.packages("gifski")
gifski::save_gif(terra::animate(summer_agg, n = 1),
                       delay = 1,
                       loop = FALSE,
                       gif_file = "summer_sst.gif")

```
# __Conclusion__

This vignette aimed to provide a general flow of operations needed to 
manipulate rasters and extract environmental variables. There are of course 
many more and alternative ways to do what I have done here, but I aimed to
demonstrate the versatility and power of the `terra` package. 
















