---
title: "Working with rasters in the terra package"
author: "Anna Stroh"
output:
  html_document: default
  pdf_document: default
date: "2024-09-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# keep code just for knitting - to be copied into console
#pandoc::with_pandoc_version(
 # version = '2.9.1',
  #rmarkdown::render('input.Rmd')
#)

```

This vignette focuses primarily on rasters, and aims to demonstrating 
fundamental raster operations needed to pre-process spatiotemporal rasters for 
further use, i.e. modelling. This vignette includes code to:

* load, 
* visualise,
* subset,
* manipulate, and 
* extract data into csv formats 

from spatiotemporal raster files using the R package `terra`. 

This script provides a workflow from loading environmental raster data and trawl
survey from online databases into R, pre-process the raster data, and extract
raster data for trawl survey data. The final product will be a data frame with
the trawl survey data and its corresponding environmental data point. 

## _Relevant packages_
```{r install packages}
### IF NOT ALREADY INSTALLED

# Raster and vector manipulation
install.packages("terra") 
install.packages("sf")

# Plotting and raster work aids
install.packages("ggplot2")
install.packages("tidyterra")
install.packages("viridis") # colour-blind colour scale

# Data wrangling
install.packages("dplyr")
install.packages("tidyr")
install.packages("lubridate")

```

## _Relevant packages_
```{r load packages}

library(terra) # raster work
library(sf) # vector work

library(ggplot2)
library(tidyterra) 
library(viridis)

library(dplyr)
library(tidyr)

```

## _Load raster data from ERDDAP servers_

Spatiotemporal grid data can be accessed within R using the `rerddap` package. 
Since the currently available ERDDAP data does not overlap with available survey
data, I will only demonstrate how to use `rerddap`. 

The loaded data is significant wave height, a value calculated from the Irish 
Shelf SWAN wave model. For the exercise we use a 2022 hindcast of wave height 
from that model.   

```{r rerddap data}
#install.packages("rerddap")
library(rerddap)

?rerddap

# What are publicly available ERDDAP servers?
AllServers <- rerddap::servers()
names(AllServers) # name and url

# Search for available data sets for a given variable on MI server
whParameter = "significant_wave_height"
urlBase = "https://erddap.marine.ie/erddap/"
whichWH <- ed_search(query = whParameter, 
                      url =  urlBase, # server of interest
                      which = "griddap") # gridded or tabulated data?

# Get info on the dataset
whInfo <- rerddap::info('IMI_IRISH_SHELF_SWAN_WAVE', 
                url =  urlBase)

# Load data into R session
#whTimes <- c("2025-01-19T00:00:00Z", "2025-01-29T00:00:00Z")
whLats <- c(51, 54.5) # Western Ireland
whLons <- c(-13, -9) 

whDat <- griddap(whInfo, 
                  longitude = whLons, 
                  latitude = whLats,
                  #time = sstTimes, # omitted for changes in data availability
                  fields = whParameter,
                  url = urlBase,
                  fmt = "nc", # netcdf file format; forces storage on disk
                  read = FALSE) # do not read into memory


file.size(whDat$summary$filename)/1e6 ## in MB

example <- rast(whDat$summary$filename)
plot(example)

rm(list = c("whDat", "example")) # optional if you have less memory space

```

## _Reading and visualising spatio-temporal rasters in terra_

For the exercise, we use wave height data that was previously requested for my 
previous research project. Please copy this link, paste into your browser and 
download the file:
https://www.dropbox.com/scl/fi/d505u992jzswppckcu11d/wh_time_sub.nc?rlkey=2h38k
jn4soovolyqxyb96hmze&st=vbvmfvab&dl=0

In `terra`, `rast()` creates a SpatRaster object. The functions takes the common 
file types as input. The GDAL driver recognises them by the filetype ending 
(i.e. filename.nc - NetCDF).

```{r read rasters}

#your_path <- ""
your_path <- getwd()
raster_file <- "wh_time_sub.nc" # include filetype (.nc) for optimised read

## Read file into R
wh_rast <- rast(paste0(your_path, "/", raster_file))

## Explore file 
wh_rast

# spatial dimensions
dim(wh_rast)
nrow(wh_rast)
ncol(wh_rast)
#crs(wh_rast)
res(wh_rast)
ext(wh_rast)

# time dimension
head(time(wh_rast))
length(time(wh_rast)) == nlyr(wh_rast) # each time stamp = one raster layer
#min(time(wh_rast))

# variable and raster layers
varnames(wh_rast) # single band raster
head(names(wh_rast)) # multilayer raster
#nlyr(wh_rast)

# raster values
#head(values(wh_rast)) # caution - requires lots in memory

```

The file contains a hindcast of significant wave height in the months November-
December 2022. `nlyr()` shows you that it is a multilayered raster (raster cube).      
`time()` shows you that each time stamp represents a raster layer. The present
raster provides the value of significant wave height every three hours between 
November-December.

`rast()` does not differentiate between spatial or spatiotemporal rasters. This
is because a SpatRaster is represented by three dimensionsional cells (rows, 
columns, layers). `terra` is efficient with large raster files because a 
SpatRaster object only stores the parameters that describe the raster file, but 
not the cell (pixel) values.  

```{r visualise rasters}
# terra method
plot(wh_rast)

# ggplot method (requires package "tidyterra")
ggplot() +
  geom_spatraster(data = wh_rast[[1:4]], # reduce nlyr for plotting
                  maxcell = ncell(wh_rast[[1:4]])) + 
  scale_fill_viridis_c(name = "wave \nheight (m)") +
  # if displaying multiple layers, use "facet_wrap( ~ lyr)"
  facet_wrap(~ lyr, # no substitute for layer names necessary - use "~lyr"
             ncol = 2, # customise layout
             nrow = 2) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 90))


```
## _Loading ICES trawl survey data_ 

```{r icesDatras}
#install.packages("icesDatras")
library(icesDatras)

# What surveys are available?
getSurveyList()

# Basic queries of standard data
#getCAdata() # age-based data 
#getHLdata() # length-based data
#getHHdata() # haul information 

# Get HH data for Irish Groundfish Survey (IE-IGFS)
HHigfs <- getHHdata(survey = "IE-IGFS", year = 2022, quarter = 4)
head(HHigfs)
names(HHigfs)

HHigfs <- HHigfs |>
  select(Survey, HaulNo, Year, Month, Day, 
         ShootLat, ShootLong) |>
  mutate(Survey = "IGFS",
         HaulID = paste0(Survey, Year, "H", HaulNo),
         SamplingDate = lubridate::make_date(Year, Month, Day))

# Make sf object 
HHigfs_sf <- HHigfs |>
  st_as_sf(coords = c("ShootLong", "ShootLat"), crs = crs(wh_rast))
head(HHigfs_sf)

```

```{r plot survey}

## Convert sf object to terra SpatVector (example sf-terra interaction)
HHigfs_vect <- vect(HHigfs_sf)

plot(wh_rast[[1]], main = "Original raster with 2022 IGFS hauls") # select L248-249
plot(HHigfs_vect, add = TRUE, col = "white", lwd = 3.5)
  
```
## _Cropping a raster to a desired spatial extent_ 

Suppose, we are only interested in the hauls completed in ICES areas 7j-g. 
To make the data extraction more efficient, we can crop the raster to the 
desired spatial range. 

To crop a raster to a custom spatial extent, raster object or spatial vector, 
use `crop()`.

A vector can be in the shape of points, lines or polygons. Spatial vectors 
consist of a single dimension with a distinct spatial feature geometry 
(i.e. polygon).

In `terra`, vectors are created using `vect()`, which creates a 
`SpatVector` object (either from scratch or from an sf object).  

```{r create SpatVector}

## Create MPA SpatVector based on extent
e <- ext(-12.35, -5, 49.9, 52.5) # vector extent for ICES areas 7 j-g
areas7jg <- vect(e) # create vector (SpatVector)
crs(areas7jg) <- 'wgs84'

areas7jg 
plot(areas7jg)

## Plot ICES areas with raster
plot(wh_rast[[1]], main = "Original raster with ICES area 7j-g cut-off")
plot(areas7jg, add = TRUE, col = "transparent", border = "white", lwd = 3.5)

## Crop wave height SpatRaster to ICES area SpatVector
ICESarea_rast <- crop(wh_rast, areas7jg)  
ICESarea_rast # produces SpatRaster

plot(ICESarea_rast[[1:4]])

## Crop haul points to ICES area SpatVector
HHigfs_coords <- HHigfs |> 
  select(HaulID, ShootLong, ShootLat) |>
  distinct() |>
  vect(geom = c("ShootLong", "ShootLat"), crs = "wgs84") # make df a SpatVector
dim(HHigfs_coords)

HHigfs_7jg <- crop(HHigfs_coords, areas7jg)  
HHigfs_7jg # produces SpatRaster

plot(HHigfs_7jg)

rm(list = c("wh_rast", "HHigfs_coords", "HHigfs_vect")) # remove original rast from environment

```

## _Subsetting and aggregating rasters_

Subsetting raster layers and time dimension in R is done using double square
brackets `[[ ]]`.

Subsetting layers is usually aided by using numerical indices that match the 
order of layers. The same goes for subsetting the time dimension.  

```{r Subset terra rasters}

## Subset SpatRaster layers

# to single layer
subset1 <- ICESarea_rast[[1]] # [[ layer index ]]

# custom number of layer
subset2 <- ICESarea_rast[[c(1, 3:6, 9)]] # [[ c(layer indexes) ]]

# or using function
subset3 <- subset(ICESarea_rast, 1)

```

Before we do a meaningful subset of the time dimension, we need to aggregate 
the raster file. 

`terra` offers temporal aggregation and spatial aggregation:

* Temporal aggregation: aggregates raster values over chosen time interval,
                        creates new raster with lower temporal resolution
    + `tapp()`

* Spatial aggregation: aggregates groups of cells to create larger cells,
                       creates new raster with lower spatial resolution
    + `aggregate()`

Here, we are working only with temporal aggregation. `tapp()` accesses raster 
layers indexed by a time index (i.e. years), and aggregates 
these layers over a given function (i.e. mean).

```{r Aggregate}

## Aggregate hourly time stamps to daily mean
D_mean_wh <- tapp(ICESarea_rast, "days", fun = mean)
D_mean_wh # note change in layer names and time dimension

plot(D_mean_wh[[1:6]]) 

```

Now, we can further pre-process the raster by subsetting the raster time 
dimension to the dates of the survey hauls. Just as before, we create an index 
that is matching the values in the time dimension. Be careful in this step not
to overwrite the time dimension of the original raster if you still need it. 

```{r TimeDim Subset}

## Subset raster time dimension
head(time(D_mean_wh))
class(time(D_mean_wh))

haul_dates <- lubridate::ymd(HHigfs_sf$SamplingDate) # matching index
#class(haul_dates)

wh_IGFS_sub <- D_mean_wh[[ time(D_mean_wh) %in% haul_dates ]] # subset time and assign to new raster

rm(list = c("D_mean_wh")) # optional

```

# Part 4: __Raster extraction__

The last part of this vignette demonstrates how to extract raster values for a 
given spatial geometry (point in this case). This is particularly useful for
linking raster-based environmental data to biological data at sampling locations 
(i.e. stations). 

Here, we aim to link mean daily wave height to the IGFS sampling stations in 
ICES areas 7j-g. 

Raster data for a given object is extracted using the `extract()` function.

To bind the extracted data, the most straightforward way is to include the 
argument `bind=TRUE` which cbinds the extracted raster values to the 
`SpatVector` of the sampling stations. 

We can then transform the `SpatVector` with the extracted value into a 
data frame to further shape it into a suitable format.  

``` {r extract data points}

## Extract wave height data
# reduce to coordinates make SpatVector from dataframe
wh_extr <- terra::extract(wh_IGFS_sub, # SpatRaster
                          HHigfs_7jg, # SpatVector
                           # returns value from the cell the coords fall into
                          method = "simple",
                           # binds raster values to vector
                          bind = TRUE) # do not return cell ID

head(wh_extr) 

# make dataframe
wh_extr_df <- terra::as.data.frame(wh_extr) # wide format
head(wh_extr_df)

## Reshape and clean-up

#install.packages("reshape2")
# wide to long
wh_extr_melt <- reshape2::melt(wh_extr_df, 
                                id = c("HaulID"))
head(wh_extr_melt)

station_wh <- wh_extr_melt |>
  
  # rename layers
  dplyr::rename(SamplingDate = variable,
                DAvWh = value) |>
  
  # clean up year variable
  mutate(SamplingDate = lubridate::ymd(gsub("d_", "", SamplingDate))) |>
  
  # merge in station sampling date and locations
  merge(HHigfs) |>
  
  # change order of columns
  relocate(SamplingDate, .before = SamplingDate)

head(station_wh) # FINAL DATA FRAME READY FOR EXPORT


## Plot final output
st_vec <- vect(station_wh, geom = c("ShootLong", "ShootLat"), crs = "wgs84")
head(st_vec)
st_vec

ggplot() +
  geom_spatvector(data = st_vec, aes(colour = DAvWh)) + # plot SpatVector
  scale_colour_viridis_c() +
  geom_spatvector(data = areas7jg, fill = NA, colour = "black") +
  labs(colour = "Daily\nAverage\nWaveheight (m)") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 90))



```
# Extra: __Raster animation__

This part is not compulsory for raster analysis, but demonstrates a fun way to
visualise raster data. `terra` has an in-built function to sequentially plot 
raster layers, which can then be exported as a video or .gif file: an animation 
of the raster. 

I am intentionally using the aggregated raster with regards to rendering 
processes. 

```{r raster animation}

## Make animation
terra::animate(wh_IGFS_sub, n = 1)

## Save animation as gif
#install.packages("gifski")
gifski::save_gif(terra::animate(wh_IGFS_sub, n = 1),
                       delay = 1,
                       loop = FALSE,
                      #gif_file = paste0(your_path, "/", "wh_ICES7jg.gif"),
                       gif_file = "wh_ICES7jg.gif")

```
# __Conclusion__

This vignette aimed to provide a general flow of operations needed to 
manipulate rasters and extract environmental variables. There are of course 
many more and alternative ways to do what I have done here, but I aimed to
demonstrate the versatility and power of the `terra` package. 
















